The `zoon` R package for reproducible and shareable species distribution modelling

### 2016-12-19 ms.Rmd
merging software comparison in introduction and tweaking modules and chains etc. sections

Like many areas of quantitative science species distribution modelling (SDM) has grown rapidly, fueled by the availability of diverse modelling software and data [@Barbosa2015].
Over the last 20 years, SDM has developed into a very large research community and vast literature, and is now one of “the most widely-reviewed topics in the ecological literature” [@Araujo2012].
As in other computational sub-disciplines, most SDM software focusses on analytical tasks such as constructing models, rather than enabling scientists to share software, and to produce reproducible and modifiable analyses that are essential to the scientific enterprise.

@@

This paper introduces version 0.4-22 of the zoon R package.
Firstly, we describe the modular structure of zoon workflows and how they can be constructed, shared, reproduced and modified.
We then illustrate how these concepts enable better SDM research by reproducing a published SDM analysis; converting a recently proposed method into a zoon module; and extending a published methodological comparison.

### Reproducible software projects

Most of the SDM users replying to a survey in 2015 [@Ahmed2015] were using either MaxEnt [@Phillips2006] or R [@R] as the first choice software for their analyses (also see @Touchon2016). 
The software market is diverse, however, including a variety of software developed specifically for SDM (e.g. MaxEnt [@Phillips2006], the BioMod and BioMod2 [@Thuiller2009] and dismo [@dismo] packages for R, SDMtoolbox [@Brown2014], openModeller [@deSouza2011], BioEnsembles [@Diniz2009], ModEco [@Guo2010]) and software that has been appropriated for SDM studies, such as statistical software (WinBugs and OpenBugs [@Lunn2000]) and other, more general ecological software  (Domain [@Carpenter1993], Canoco [@TerBraak2002]).

Of the SDM software for multi model comparisons (e.g. OpenModeller, BioEnsembles, ModEco), BioMod (as an open-source R package) has the greatest potential to be modified and re-appropriated.
In essence, BioMod is an SDM environment accessed from R, rather than necessarily being a SDM tool kit for R.
The R package called sdm [@Naimi2016] implements a variety of analytical functionality in a simplifying framework.
There is scope for incorporating new techniques but not necessarily a straight forward system for sharing those contributions.
The BIOVEL workflow environment [@DeGiovanni2015] is an exception in terms of reproducibility, sharing many of the general principles of `zoon`.
Adoption of workflow environments (e.g. Kepler, Vis-Trails, Taverna) is low however, presumably due to the users’ unfamiliarity, and the requirement of embedding R within a workflow environment rather than embedding a workflow system within R.

Examples of subject specific frameworks are found within the Bioinformatics community, for example the Bioconductor infrastructure that uses R (@Gentleman2004, `bioconductor.org`), and BioJS, a repository for JavaScript components (@Corpas2014, `biojs.net`).
Both software aim to increase the compatibility and sharing of computational tools in order to increase the reuse and robustness of tools.
In terms of robustness, the CHASTE project for cardiac modelling (@Mirams2013, `cs.ox.ac.uk/chaste`) has developed into a model repository that offers community benchmarks for the last 50 years of cardiac modelling [@Cooper2016].
These frameworks increase the access, usability and robustness of scientific research.
Of course, R users could point to the R environment itself, but the stringent requirements of R packages are likely to prevent much of the SDM community from sharing code in a discoverable or usable format.

### The Zoon Project

The Zoon Project takes a different approach, instead defining only the core architecture of workflows, and interfacing with an open repository of community-contributed software modules.

Users can create new modules for any modelling method and integrate it into another analysis. 
Similarly to programming languages such as R and Python, facilitating wider development of software will enable the whole research community to develop the software that it requires, rather than rely on the developers of the software they use to do it for them.

Unlike software packages for programming languages however, The Zoon Project enforces a modular structure that ensures interoperability between contributed software modules.
This also helps to lower the bar for creating modules (each module is simply an R function with a set of required inputs and outputs) and facilitates automated testing of modules to ensure software quality.

@@

@@

####

@@

@@

@@

The complete workflow, with multiple modules defined with the `Chain` function and parallel analyses being defined with the `list` or `replicate` function can be visualised by calling `plot` with the workflow object as it's argument.
This displays the names of all modules and the broader structure of the workflow as well as displaying which modules are not available in the online repository (Figure ).

```{r plotworkflow, eval = TRUE, echo = FALSE, fig.height = 4, fig.width = 9, dpi = 300}

work <- workflow(occurrence = UKAnophelesPlumbeus,
                 covariate  = UKAir,
                 process    = Chain(Background(n = 70), Crossvalidate(k = 5)),
                 model      = list(LogisticRegression, RandomForest, QuickGRaF),
                 output     = PerformanceMeasures)

plot(work)
```

@@

A `zoonWorkflow` objectStructure of workflow objects:

>>>>

Like many areas of quantitative science species distribution modelling (SDM) has grown rapidly, fueled by the availability of diverse modelling software and data [@Barbosa2015].
Over the last 20 years, SDM has developed into a very large research community and vast literature.
SDM is now one of “the most widely-reviewed topics in the ecological literature” [@Araujo2012] and the growth of this literature is accelerating [@Barbosa2015].

The SDM software market is similarly diverse [@Ahmed2015].
It comprises many software packages developed specifically for SDM such as the BioMod, BioMod2 [@Thuiller2009], dismo [@dismo] and sdm [@Naimi2016] R packages, the SDMtoolbox python package [@Brown2014] and GUI applications including MaxEnt [@Phillips2006], openModeller [@deSouza2011], BioEnsembles [@Diniz2009] and ModEco [@Guo2010].
A range of other software has also been appropriated for SDM studies, such as statistical software including WinBugs and OpenBugs [@Lunn2000], and more general ecological software such as Domain [@Carpenter1993], Canoco [@TerBraak2002].
However most SDM users replying to a survey in 2015 [@Ahmed2015] used either MaxEnt [@Phillips2006] or R [@R] as the first choice software for their analyses.
As in other computational sub-disciplines, most SDM software focusses on analytical tasks such as constructing models, rather than enabling scientists to share software, and to produce reproducible and modifiable analyses that are essential to the scientific enterprise.

@@

### Comparison with existing software

The modular nature of zoon workflows will be familiar to users of other SDM R packages, such as BioMod2 [@Thuiller2009] and sdm [@Naimi2016].
Whilst BioMod2 modules and sdm modules are created by the developers of those packages and contained in the software itself,zoon's modules are drawn from an open repository that any SDM user can contribute to.
zoon makes module writing easy for SDM users who are not expert R developers, and provides functionality to help document, test and upload modules to the repository.
zoon workflows are also designed as reproducible research objects which can be shared, inspected, re-run and modified by swapping in new modules.

This focus on reproducibile and modifiable workflows shares similarities with general-purpose workflow systems such as Kepler [@], Vis-Trails [@], Taverna [@].
The Taverna platform has recently been extended for biodiversity research via the BIOVEL system [@DeGiovanni2015] and provides some SDM functionality via its ENM Components [@DeGiovanni2016].
However adoption of these workflow environments in the SDM research community is low.
Presumably this is due to SDM users’ unfamiliarity with workflow software and the limited number of methods already implemented.
This hurdle is created by requiring users to embed R code within a workflow software, rather than embedding a workflow system within R.

By constrast with existing software, zoon has been custom-built to improve reproducibility of SDM research in R.
By providing a single, extensible access point to various SDM software, it is designed to facilitate a more connected and effective SDM research community.

@@

This paper introduces version 0.4-22 of the zoon R package.
We describe the modular structure of zoon workflows and how they can be constructed, shared, reproduced and modified.
We then illustrate how these concepts enable better SDM research by reproducing a published SDM analysis; converting a recently proposed method into a zoon module; and extending a published methodological comparison.

@@

@@

Users can query the modules hosted on the repository from their R session using the function `GetModuleList()`, and pull up the documentation for any module in the repository with `ModuleHelp()`.

@@

Many SDM analyses apply more than one method for each of these steps.
zoon enables more complex workflows by enabling users to pass multiple modules of each type via the List, Chain and Replicate functions.

@@

The following workflow demonstrates use of the `Chain`s and `list`s to fit three different SDMs to a presence-background dataset for a UK mosquito species, after standardising the covariates. The results of the three models are then returned in Shiny app allowing the user to interactively explore the occurrence and covariate data, model summaries and prediction maps.

```{r plotworkflow, eval = TRUE, echo = TRUE, fig.height = 4, fig.width = 9, dpi = 300}
work <- workflow(occurrence = UKAnophelesPlumbeus,
                 covariate  = UKBioclim,
                 process    = Chain(Background(n = 500),
                                    StandardiseCov),
                 model      = list(MaxEnt,
                                   GBM,
                                   RandomForest),
                 output     = Appify)
```

@@

A `zoonWorkflow` object
Structure of workflow objects:
